---
title: "Chapter 5: SRQ R plots"
author: "Joseph Ho"
date: "2023-06-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Smooth check functions and Systematic Regression Quantiles (SRQ)
Reference: Saleh R.A., Saleh A.K.Md.E. Solution to the Non-Monotonicity and Crossing Problems in Quantile Regression. arXiv [Preprint] 2021. Version 2. [URL](https://arxiv.org/abs/2111.04805)

Original check function for quantile regression is
$$\rho_\tau(u) = u(\tau - \mathbb{I}_{\{u < 0\}}), \qquad u \in \mathbb{R}, \tau \in [0,1]$$
It is piecewise linear but not differentiable at $u=0$. The (piecewise) derivative of $\rho_\tau$ is
$$\rho'_\tau(u) = \begin{cases} \tau - 1, & u < 0\\
\tau, & u \geq 0\end{cases}$$
The alternative proposed in the article (revised by the author) is to approximate this derivative by
$$f(x|\tau, c) := \frac{1}{2}\tanh(cx) + (\tau - \frac{1}{2}), \qquad \tau \in [0,1], c \geq 1$$
where $c \geq 1$ is a parameter to be specified (choose a large $c$ since $\tanh(cx) \to \text{sgn}(x)$ pointwise as $c \to \infty$). Then the continuous check function is
$$F(x|\tau,c) = \begin{cases}
\frac{1}{2c} \log(\cosh(cx)) + (\tau-\frac{1}{2})x + \frac{1}{2c} \log 2, & \tau = 0 \text{ or } \tau = 1\\
\frac{1}{2c} \log(\cosh(cx)) + (\tau-\frac{1}{2})x + \frac{1}{2c} \log 2 + \frac{1}{2c}[\log(\sqrt{\tau(1-\tau)})-(\tau-\frac{1}{2})\log(\frac{1-\tau}{\tau})], & 0 < \tau < 1
\end{cases}$$

Now we try to replicate the plots in this paper in Section IV.C:
```{r anscombe data 1 quantreg median}
library(quantreg)
plot(anscombe$x1, anscombe$y1)
fit1 <- rq(anscombe$y1 ~ anscombe$x1, tau = 0.5) # median regression
summary(fit1)
abline(fit1, col="blue") # median regression line (blue solid line)
abline(lm(anscombe$y1 ~ anscombe$x1), lty = 2, col = "red") # linear regression line (red dotted line)
```



We now implement the method of SRQ on the El Nino data.

Source of data: El Nino. UCI Machine Learning Repository. 1999. \url{https://doi.org/10.24432/C5WG62}

We use the observations in 1997 (where El-Nino occured) that are cleaned (and omitted observations with missing data).

```{r El-nino data, echo=FALSE}
library(readr)
elnino <- 
  read_csv("./elnino_clean_final_1997.csv") # import cleaned data
#head(elnino) # 14776 observations

# define variables
latitude <- elnino$Latitude
longitude <- elnino$Longitude
zonalwinds <- elnino$Zonal_Winds
meridionalwinds <- elnino$Meridional_Winds
humidity <- elnino$Humidity
airtemp <- elnino$Air_Temp
seasurfacetemp <- elnino$Sea_Surface_Temp
```


```{r El-nino data quantreg quantile functions, echo=FALSE,results='hide',fig.keep='all'}
library(quantreg)
library(viridis)
plot(airtemp, seasurfacetemp, main="Quantile functions of
    El Nino data 1997", cex=0.5)
n <- 10 # number of quantile functions (lines)
for (i in 1:(n-1)){
  elnino_fit_quant <- rq(seasurfacetemp ~ airtemp, tau = i/n)
  abline(elnino_fit_quant, col=viridis(n-1)[i]) # quantile regression line (solid lines)
}
#abline(lm(seasurfacetemp ~ airtemp), lty = 2, col = "red") # linear regression line (red dotted line)
```
From the plot, we see quantile crossing between the upper quantile functions at airtemp=29.8.

```{r elninodata SRQ single covariate}
#library(quantreg)
library(viridis)
y <- c(seasurfacetemp)
x <- matrix(airtemp)
X <- cbind(rep(1,nrow(x)),x)

# plots
plot(x, y, xlab = "airtemp", ylab = "seasurfacetemp", main="SRQ Quantile functions of
    El Nino data 1997", cex=0.5)

# quantile regression using SRQ on discrete set of taus
n <- 10 # number of quantile functions
for (i in 1:(n-1)){
  betas_SRQ <- srq(X,y,tau = i/n)
  abline(a=betas_SRQ[1], b=betas_SRQ[2], col = viridis(n-1)[i]) # SRQ quantile regression line
}
```
The SRQ quantile functions still cross at airtemp=29.8, which implies SRQ cannot solve the quantile crossing problem.